### 2014-10-24 intro.tex
Intro modified

Many real-world web services --- such as Amazon, Facebook, Google, Twitter ---
replicate application state and logic across multiple \emph{replicas} within
and across data centers. Replication is intended not only to improve
application throughput and reduce user perceived latency, but also to tolerate
partial failures without compromising the service availability. Traditionally
programmers have relied on \emph{strong consistency} guarantees such as
Linerarizability~\cite{Herlihy1990} or serializability~\cite{Serializability}
in order to build correct applications. While strong consistency is simple to
reason about, the CAP theorem~\cite{Brewer2000,Gilbert} makes it impossible for
a system to achieve strong consistency, tolerate network partitions and remain
highly available. Thus, modern web services, which aim to provide an "always
on" experience, overwhelmingly embrace availability and partition tolerance
over strong consistency. Several \emph{weak consistency} models such as
eventual consistency, causal consistency, session guarantees, and time line
consistency have been proposed.

Under weak consistency, the programmer needs to be aware of concurrent
conflicting updates, and has to pay careful attention to avoid inconsistencies
such as negative balances in a bank account, and an item appearing in the
shopping cart after it has been removed~\cite{Dynamo}. Often times, the
inconsistency leaks from the application and is witnessed by the user. To
address this issue, many commercial~\cite{} and research~\cite{} systems
propose guarantees that are weaker than strong consistency, and hence need only
to contact a subset of replicas for discharging the operations. With these
alternatives, it is up to the programmer to decide whether a weaker consistency
level is sufficient for a preserving the application-level consistency
properties. Understandably, this is an error prone process as the user not only
has to understand the intricate behavior of the application, but also the
semantics of the underlying store consistency levels, which are often presented
informally~\cite{}.

@@

Moreover, weak consistency hinders compositional reasoning about programs.
While the application might be naturally expressed as maps, trees, counters,
queues, or graphs, geo-distributed stores are typically only provide a minimal
set of data types with in-built conflict resolution strategy such as
last-writer-wins (LWW) register, counters and sets~\cite{Cassandra,DynamoDB}.
While traditional database systems enable composability through transactions,
the state-of-the-art geo-distributed stores lack unrestricted transactional
access to the data. Indeed, traditional database systems do offer a variety of
transaction isolation levels~\cite{}, each with a particular runtime overhead,
leaving the choice of correct consistency level to the programmer. Thus, the
programmer working with geo-distributed stores has but to coerce the
application state to the data store capabilities.

To address these issues, we introduce Quelea, a declarative programming model
for eventually consistent stores. The key novelty of Quelea is a \emph{contract
language} to express the fine-grained application-level consistency properties.
The programmer uses the contract language to describe the set of legal
executions allowed over the replicated data types. The contracts are
constructed using primitive relations such as \emph{visibility} and
\emph{session order} and a carefully chosen set of logical operators. The
executions descriptions are similar to the high-level axiomatic description of
relaxed memory models~\cite{Demange2013}. A contract enforcement system
logically analyses the contracts using an SMT solver~\cite{} and automatically
maps each operation to one of the store consistency levels. In addition, the
system statically reports an error if the given application-level consistency
property cannot be satisfied by the store. The paper makes the following
contributions:

>>>>

Many real-world web applications --- such as Amazon, Facebook, Google, Twitter
--- replicate application state and logic across multiple \emph{replicas} within
and across data centers. Replication is intended not only to improve application
throughput and reduce user perceived latency, but also to tolerate partial
failures without compromising the service availability. Traditionally
programmers have relied on \emph{strong consistency} guarantees, such as
Linerizability~\cite{Herlihy1990} or serializability~\cite{Serializability},
provided by the underlying data store in order to build correct applications.
While strong consistency is simple to reason about, the CAP
theorem~\cite{Brewer2000,Gilbert} makes it impossible for a replicated data
store to achieve strong consistency, while tolerating network partitions and
remaining highly available. Thus, modern web applications, which aim to provide
an "always on" experience, overwhelmingly embrace availability and partition
tolerance at the expense of strong consistency.

Nevetheless, to avoid exposing end user to absurdities such as negative balances
in a bank account, or an item appearing in the shopping cart after it had been
removed~\cite{Dynamo}, web applications \emph{do} require some level of
consistency guarantees from the underlying data store. Accordingly, many
commercial~\cite{} and research~\cite{} systems provide guarantees that are
weaker than strong consistency, such as eventual consistency, causal
consistency, session guarantees, and time line consistency. Each such \emph{weak
consistency} guarantee imposes certain amount of penalty on the availability
(hence, latency and throughput) of the system depending on the extent of
inter-replica coordination required to ensure the guaranteed level of
consistency. It is therefore incumbent on the application programmer to
carefully reason about concurrent conflicting updates, and select the
appropriate store-level weak consistency guarantee sufficient for preserving
application-level integrity properties, while minimizing the adverse impact on
the availability of the application. Understandably, this is an error prone
process as the user not only has to understand the intricate behavior of the
application, but also the semantics of the underlying store consistency levels,
which are often presented informally~\cite{}.

@@

Moreover, weak consistency hinders compositional reasoning about programs.
While applications are usually built around rich abstract data types, such as
dictionaries, counters, queues, and graphs, geo-distributed stores typically
only provide a minimal set of data types with in-built conflict resolution
strategy such as last-writer-wins (LWW) register, counters and
sets~\cite{Cassandra,DynamoDB}.  While traditional database systems enable
composability through transactions, the state-of-the-art geo-distributed stores
lack unrestricted transactional access to the data. Indeed, traditional database
systems do offer a variety of transaction isolation levels~\cite{}, each with a
particular runtime overhead, leaving the choice of correct consistency level to
the programmer. Thus, the programmer working with geo-distributed stores has but
to coerce the application state to the data store capabilities.

To address these issues, we introduce Quelea, a declarative programming system
for eventually consistent data stores. Quelea's programming model is built
around operation-based Commutative Replicated Data Types (CRDTs)
~\cite{shapiroCRDT}. The state of an abstract data type in Quelea is modeled as
a replicated \emph{append-only} log storing the history of all available
\emph{update} operations\footnote{Available at each replica. Different replicas
may have different logs as not all updates would have propagated to every
replica.} performed on the data type . For instance, the state of a replicated
counter ADT is stored on all replicas as a log of \emph{increment} and
\emph{decrement} operations performed on the counter. An operation that indends
to read the current value the counter at a replica only has to \emph{fold} over
the history log at that replica. Our key observation is that under this model,
consistency requirements of a replicated data type operation (eg:
\emph{decrement}) can be expressed effectively in terms of assertions on
\emph{membership} property of logs at various replicas. Quelea makes use of this
observation to introduce a simple \emph{contract language} that is nevertheless
capable of expressing fine-grained application-level consistency properties.
The contract language is made up of a couple of primitive relations, namely
\emph{visibility} and \emph{session order}, and a carefully choosen set of
logical operators. Quelea expects application programmer to use the contract
language to describe the set of legal executions allowed over the replicated
data types.  
% The executions descriptions are similar to the high-level axiomatic
% description of relaxed memory models~\cite{Demange2013}.
Subsequently, a contract enforcement system logically analyses contracts using
an SMT solver~\cite{z3} and automatically tags each operation with the weakest
store-level consistency guarantee that is required to satisfy application-level
consistency requirements. In addition, the system statically reports an error if
the given application-level consistency property cannot be satisfied by the
store.

The paper makes the following contributions:
